---
title: "Catchments (Q + T joined to COMID, stream order >= 4)"
output: html_document
editor_options:
  chunk_output_type: console
---

STEP 4 of the pipeline (runs after Watertemp.Rmd).

Purpose
-------
Build the NHDPlus HR routing layer for the AOI (flowlines + catchments),
snap every USGS site that appears in the Q+T table to a COMID, and filter
everything to stream order >= 4 in one pass. No duplicated full-basin
layers are persisted.

Final outputs (single version of each — these ARE the SO>=4 layers):
  - `nhd_dir/flowlines.gpkg`          : NHDPlus HR flowlines, SO >= 4, clipped to AOI
  - `nhd_dir/catchments.gpkg`         : matching catchment polygons (same COMIDs)
  - `daily_drivers_dir/q_t_comid.parquet` : Q + T daily obs joined to COMID, SO >= 4
  - `shape_dir/sites_with_comid.csv`  : site_no -> comid lookup for provenance

```{r setup}
source("SRC/Functions_Paths.R")

library(sf)
library(dplyr)
library(arrow)
library(dataRetrieval)
library(nhdplusTools)

sf::sf_use_s2(FALSE)
```

```{r aoi}
# AOI from Boundaries.py
aoi <- st_read(basin_shp, quiet = TRUE) %>%
  st_transform(4326) %>%
  st_make_valid()
```

```{r nhd-hr-download}
# Illinois basin = HU4 0712 + 0713
hu4s        <- c("0712", "0713")
hr_data_dir <- file.path(nhd_dir, "hr_raw")
hr_gpkg     <- file.path(nhd_dir, "nhdplushr.gpkg")

dir.create(hr_data_dir, recursive = TRUE, showWarnings = FALSE)

download_nhdplushr(hr_data_dir, hu4s)

hr <- get_nhdplushr(
  hr_data_dir,
  layers   = c("NHDFlowline", "NHDPlusCatchment"),
  out_gpkg = hr_gpkg
)
```

```{r flowlines-catchments}
# HR flowlines clipped to AOI, renamed to lowercase to match the rest of the pipeline
flowlines <- hr$NHDFlowline %>%
  st_as_sf() %>%
  st_transform(4326) %>%
  rename(
    comid      = COMID,
    reachcode  = REACHCODE,
    gnis_name  = GNIS_Name,
    lengthkm   = LENGTHKM,
    areasqkm   = TotDASqKM,
    slope      = Slope,
    streamorde = StreamOrde,
    streamcalc = StreamCalc,
    hydroseq   = Hydroseq,
    levelpathi = LevelPathI,
    fromnode   = FromNode,
    tonode     = ToNode,
    terminalfl = TerminalFl,
    startflag  = StartFlag,
    divergence = Divergence
  ) %>%
  select(comid, reachcode, gnis_name, lengthkm, areasqkm, slope,
         streamorde, streamcalc, hydroseq, levelpathi,
         fromnode, tonode, terminalfl, startflag, divergence) %>%
  st_intersection(aoi)

catchments <- hr$NHDPlusCatchment %>%
  st_as_sf(sf_column_name = "geom") %>%
  st_transform(4326) %>%
  st_make_valid() %>%
  mutate(comid = as.character(FEATUREID)) %>%
  rename(areasqkm = AreaSqKM) %>%
  select(comid, areasqkm) %>%
  st_intersection(aoi)
```

```{r orphan-check}
orphans <- flowlines %>%
  filter(is.na(fromnode) | is.na(tonode)) %>%
  pull(comid)

if (length(orphans) > 0) {
  warning("Orphan COMIDs found: ", paste(orphans, collapse = ", "))
} else {
  message("All COMIDs routable (no orphan edges).")
}
```

```{r load-qt-and-site-coords}
q_t <- read_parquet(file.path(daily_drivers_dir, "q_t.parquet")) %>%
  mutate(Date = as.Date(Date))

site_info <- read_waterdata_monitoring_location(
  monitoring_location_id = paste0("USGS-", unique(q_t$site_no)),
  skipGeometry           = FALSE
)

site_coords <- site_info %>%
  filter(!is.na(geometry)) %>%
  st_as_sf() %>%
  st_transform(4326) %>%
  mutate(site_no = gsub("^USGS-", "", monitoring_location_id))
```

```{r snap-sites-to-comid}
# Primary snap: nearest flowline with VAA measures (uses full network so small
# tributaries don't get misrouted to a mainstem reach).
vaa <- get_vaa() %>% select(comid, frommeas, tomeas)

flowlines_idx <- flowlines %>%
  left_join(vaa, by = "comid") %>%
  rename(FromMeas = frommeas, ToMeas = tomeas)

site_matches <- get_flowline_index(flowlines_idx, site_coords, search_radius = 100)

sites_with_comid <- site_coords %>% mutate(comid = site_matches$COMID)

# Fallback: nearest catchment polygon for any site that didn't snap
unsnapped <- sites_with_comid %>% filter(is.na(comid))
if (nrow(unsnapped) > 0) {
  message(nrow(unsnapped), " site(s) rescued via nearest-catchment fallback.")
  rescued <- st_join(
    st_transform(unsnapped %>% select(site_no), 5070),
    st_transform(catchments %>% select(comid), 5070),
    join = st_nearest_feature
  ) %>%
    st_drop_geometry() %>%
    distinct(site_no, comid)

  sites_with_comid <- sites_with_comid %>%
    st_drop_geometry() %>%
    select(site_no, comid) %>%
    rows_update(rescued, by = "site_no")
} else {
  sites_with_comid <- sites_with_comid %>%
    st_drop_geometry() %>%
    select(site_no, comid)
}

write.csv(
  sites_with_comid,
  file.path(shape_dir, "sites_with_comid.csv"),
  row.names = FALSE
)
```

```{r join-and-filter-so4}
# Attach stream order onto Q+T via COMID, then filter Q+T and hydro layers
# to stream order >= 4 in one consistent pass.
flow_attrs <- flowlines %>%
  st_drop_geometry() %>%
  mutate(comid = as.character(comid)) %>%
  select(comid, streamorde)

q_t_comid <- q_t %>%
  left_join(sites_with_comid, by = "site_no") %>%
  filter(!is.na(comid)) %>%
  mutate(comid = as.character(comid)) %>%
  left_join(flow_attrs, by = "comid") %>%
  filter(!is.na(streamorde), streamorde >= 4)

flowlines <- flowlines %>% filter(!is.na(streamorde), streamorde >= 4)

keep_comids <- unique(as.character(flowlines$comid))
catchments  <- catchments %>%
  mutate(comid = as.character(comid)) %>%
  filter(comid %in% keep_comids)
```

```{r write-final-outputs}
flowline_out  <- file.path(nhd_dir, "flowlines.gpkg")
catchment_out <- file.path(nhd_dir, "catchments.gpkg")
q_t_comid_out <- file.path(daily_drivers_dir, "q_t_comid.parquet")

for (f in c(flowline_out, catchment_out, q_t_comid_out)) {
  if (file.exists(f)) file.remove(f)
}

st_write(flowlines,  flowline_out,  quiet = TRUE)
st_write(catchments, catchment_out, quiet = TRUE)
write_parquet(q_t_comid, q_t_comid_out)

cat("\nFinal Q+T+COMID (stream order >= 4):\n")
print(q_t_comid %>% summarise(
  n_rows          = n(),
  n_sites         = n_distinct(site_no),
  n_comids        = n_distinct(comid),
  min_date        = min(Date, na.rm = TRUE),
  max_date        = max(Date, na.rm = TRUE),
  pct_rows_with_Q = mean(!is.na(q_m3s)),
  pct_rows_with_T = mean(!is.na(water_temp_C))
))
```

The following is to assess  

```{r}
# One row per site with assigned COMID + stream order
site_stream_order <- q_t %>%
  distinct(site_no) %>%
  left_join(sites_with_comid, by = "site_no") %>%
  filter(!is.na(comid)) %>%
  mutate(comid = as.character(comid)) %>%
  left_join(flow_attrs, by = "comid") %>%
  distinct(site_no, comid, streamorde)

# Quick diagnostic table
site_stream_order %>%
  count(streamorde, name = "n_sites") %>%
  arrange(streamorde) %>%
  print()
```

```{r}
# Rebuild an sf version of sites with stream order attached
site_points_map <- site_coords %>%
  left_join(site_stream_order, by = "site_no")

plot(st_geometry(flowlines),
     col = "grey70",
     lwd = 0.4,
     main = "USGS sites snapped to NHDPlus HR flowlines")

plot(st_geometry(site_points_map),
     col = "red",
     pch = 16,
     cex = 0.5,
     add = TRUE)
```

```{r}
flowlines_map <- flowlines %>%
  mutate(
    so_group = case_when(
      is.na(streamorde) ~ "NA",
      streamorde == 1 ~ "1",
      streamorde == 2 ~ "2",
      streamorde == 3 ~ "3",
      streamorde >= 4 ~ "4+"
    )
  )

site_points_map <- site_coords %>%
  left_join(site_stream_order, by = "site_no") %>%
  mutate(
    so_group = case_when(
      is.na(streamorde) ~ "NA",
      streamorde == 1 ~ "1",
      streamorde == 2 ~ "2",
      streamorde == 3 ~ "3",
      streamorde >= 4 ~ "4+"
    )
  )

flow_cols <- c("1" = "#9ecae1", "2" = "#6baed6", "3" = "#3182bd", "4+" = "#08519c", "NA" = "grey80")
site_cols <- c("1" = "red", "2" = "orange", "3" = "gold", "4+" = "darkgreen", "NA" = "black")

plot(st_geometry(flowlines_map),
     col = flow_cols[flowlines_map$so_group],
     lwd = 0.6,
     main = "Flowlines and USGS sites by assigned stream order")

plot(st_geometry(site_points_map),
     col = site_cols[site_points_map$so_group],
     pch = 16,
     cex = 0.7,
     add = TRUE)

legend("bottomleft",
       legend = c("SO 1", "SO 2", "SO 3", "SO 4+"),
       col = c("red", "orange", "gold", "darkgreen"),
       pch = 16,
       bty = "n")
```

